\large Testing cosmic isotropy and homogeneity \\ with the \textsl{Quaia} all-sky spectroscopic sample of quasars

### 2023-07-03 paper/main.tex
Updates from Overleaf

% ## to-do items
% --------------
% - test the cmb dipole; cite Secrest et al. NO -- ABBY's PAPER -- BUT DISCUSS THIS POINT.
% - color choices?
% - explain, justify, or change how DD, DR, RD, and RR are counted -- the asymmetry between the tree and the sample compared to the tree.
% - discuss more philosophically why alms are sensitive tests of anisotropy
% - actually measure the fractal dimension
% - make some figure that quantitatively checks Gaia calibration
% - put the expectation hatN_DD onto the non-divided cumulative number plot
% - put in all missing references
% - address all CITE, XXX, YYY, ZZZ, HOGG, KSF in the text
% - make sure the figure axis labels and figure captions use precisely the math symbols used in the text; maybe also turn on TeX in matplotlib?

@@

\title{\large%
Testing cosmic isotropy and homogeneity \\ with the all-sky \textsl{Quaia} sample of quasars}
\author[0000-0003-2866-9403]{David W. Hogg}
\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University}
\affil{Max-Planck-Institut f\"ur Astronomie}
\affil{Flatiron Institute, a division of the Simons Foundation}

@@

\begin{abstract}\noindent
All cosmological models in general relativity are built on the assumption of large-scale isotropy and homogeneity.
These assumptions are testable.
Tests of cosmic isotropy and homogeneity are also strong tests of the quality of data or catalogs and their calibration.
Test precision increases as observations cover more solid angle, become more sensitive to distant objects, and are made with more uniform data.
The ESA \textsl{Gaia} Mission is primarily concerned with stars in the Milky Way, but it also observes millions of quasars, all sky (modulo Galactic extinction), to redshifts of around $4$.
It has produced the largest-comoving-volume public quasar catalog to date.
Here we use the $755\,850$ quasars in the \textsl{Quaia} $G<20.0$ all-sky quasar sample, built from \textsl{Gaia} and \textsl{unWISE} catalogs, to show that the angular distribution, redshift distribution as a function of hemisphere, amplitude of clustering as a function of sky position, and counts of pairs as a function of separation (fractal dimension), are all consistent with large-scale isotropy and homogeneity at the precision possible given the size of the catalog and the amplitude of the large-scale structure.
Roughly speaking, the results can be summarized by saying that they show the Universe to be isotropic at the sub-percent level and homogeneous in the sense of a fractal dimension of 3 to within YYY.
There is no sign of any North--South density or power asymmetry.
\abby{Is this the same hemispherical power asymmetry claimed from the CMB?}
The isotropy results can also be interpreted as a strong test of the spectrophotometric uniformity of the \textsl{Quaia} Catalog.
\end{abstract}

\keywords{%
calibration
---
catalogs
---
cosmic~isotropy
---
cosmological~principle
---
cosmology
---
large-scale~structure~of~the~Universe
---
quasars
}
\section*{}
\clearpage
\section{Introduction}\label{sec:intro}\noindent
One of the fundamental assumptions of the physical cosmological model is that the Universe is isotropic and homogeneous on large scales (CITE).
This assumption is testable.
It is worthy of test, both because it is always worth testing fundamental assumptions, and because the large-scale structure has self-similar or fractal characteristics (CITE), which are characteristic of inhomogeneous distributions.
And, even in the ascendent physical model of cosmology, there is expected to be at least some power in the initial conditions on all length scale (CITE).

@@

All that said, we know now that the Universe is not a fractal on large scales (\citealt{hogg05}; CITE OTHERS).
There are other reasons to continue to test isotropy and homogeneity however.
One is that the cosmic microwave background appears to show a power asymmetry, in which there appears to be a larger amplitude of fluctuations in the North/South Ecliptic (HOGG CHECK) Cap than there is in the South/North (\citealt{wmapanisotropy}; \citealt{planckanisotropy}).
If this asymmetry is real, it should show itself in the large-scale structure as well (\citealt{zhai}).

@@

\begin{figure}[t!]
  \begin{mdframed}
  \color{captiongray}
  \begin{center}
    \includegraphics[width=\figurewidth]{notebooks/cumulativeDD_DR.png}
  \end{center}
    \caption{\textsl{Top:} The cumulative total number $N_{DD}$ \eqref{eq:DD} of quasar--quasar pairs as a function of comoving separation $s$ for the NGC and SGC subsamples. Jackknife uncertainty estimates are shown but are smaller than the points for all points.
    \textsl{Bottom:} The same, but divided by the expectation $\hat{N}_{DD}$ \eqref{eq:DDhat} in a homogeneous universe.
    This figure demonstrates that the fractal dimension is extremely close to $3$.\label{fig:cumulative}}
  \end{mdframed}
\end{figure}
The cumulative number $N_{DD}$ of pairs up to separation $s$ is shown as a function of $s$ in the top panel of \figref{fig:cumulative}.
The top panel shows the raw counts of pairs, and the bottom panel shows the pairs relative to an expectation derived from the random catalog, which (by construction) has fractal dimension $d=3$.
Although the scaling in the top panel of \figref{fig:cumulative} looks very good, it can be biased by the finite volume within which we have observations; hence the ratios plotted in the bottom panel of \figref{fig:cumulative}.
In detail, the statistics that are plotted are cumulative pair counts obtained as follows:
We start by counting pairs within the quasar sample, between the quasar sample and the random catalog, and within the random catalog, by
\begin{align}\label{eq:DD}
    N_{DD}(s) &= \sum_n \sum_{n'\ne n} I(r_{nn'} < s) \\
    N_{DR}(s) &= f\,\sum_n \sum_{m'} I(r_{nm'} < s) \\
    N_{RD}(s) &= f\,\sum_m \sum_{n'} I(r_{mn'} < s) \\
    N_{RR}(s) &= f^2\,\sum_m \sum_{m'\ne m} I(r_{mm'} < s)\label{eq:RR} ~,
\end{align}
where the sums over $n$ or $n'$ are over quasars,
the sums over $m$ or $m'$ are over random points,
$r_{nn'}$ is the comoving distance between quasar $n$ and quasar $n'$,
$I(r<s)$ is an indicator function that returns 1 if $r<s$ and 0 otherwise,
and $f$ is a dimensionless ratio of the density of the quasar sample (on the sky) to the density of the random catalog (on the sky).
With these counts in hand, the expectation $\hat{N}_{DD}(s)$ we use for the cumulative pair counts $N_{DD}(s)$ is given by
\begin{equation}\label{eq:DDhat}
    \hat{N}_{DD}(s) = N_{DR}(s) + N_{RD}(s) - N_{RR}(s) ~,
\end{equation}
which is inspired by the currently standard correlation-function estimator (\citealt{ls}).

HOGG: \figref{fig:cumulative} strongly suggests that the fractal dimension $D$ is 3.
Our best-fit value is ... WHAT?
Get quantitative here.

@@

\begin{thebibliography}{dummy}
\bibitem[Eisenstein et al.(2005)]{baf} Eisenstein, D.~J., Zehavi, I., Hogg, D.~W., et al.\ 2005, \apj, 633, 560. doi:10.1086/466512
\bibitem[Eriksen et al.(2004)]{wmapanisotropy} Eriksen, H.~K., Hansen, F.~K., Banday, A.~J., et al.\ 2004, \apj, 605, 14. doi:10.1086/382267
\bibitem[Hoffleit \& Jaschek(1991)]{bsc}
Hoffleit, D. \& Jaschek, C.\ 1991, The Bright Star Catalog. New Haven: Yale University Observatory, 5th rev. ed.
\bibitem[Hogg et al.(2005)]{hogg05} Hogg, D.~W., Eisenstein, D.~J., Blanton, M.~R., et al.\ 2005, \apj, 624, 54. doi:10.1086/429084
\bibitem[Landy \& Szalay(1993)]{ls} Landy, S.~D. \& Szalay, A.~S.\ 1993, \apj, 412, 64. doi:10.1086/172900
\bibitem[Planck Collaboration(2014)]{planckanisotropy} Planck Collaboration, Ade, P.~A.~R., Aghanim, N., et al.\ 2014, \aap, 571, A23. doi:10.1051/0004-6361/201321534
\bibitem[Planck Collaboration(2020)]{planck18} Planck Collaboration, Aghanim, N., Akrami, Y., et al.\ 2020, \aap, 641, A6. doi:10.1051/0004-6361/201833910
\bibitem[Schlegel et al.(1998)]{sfd} Schlegel, D.~J., Finkbeiner, D.~P., \& Davis, M.\ 1998, \apj, 500, 525. doi:10.1086/305772
\bibitem[Storey-Fisher et al.(2022)]{ksf} Storey-Fisher, K, Hogg, D.~W., Eilers, A.-C., et al.\ 2022, in preparation
\bibitem[Zhai \& Blanton(2017)]{zhai} Zhai, Z. \& Blanton, M.~R.\ 2017, \apj, 850, 41. doi:10.3847/1538-4357/aa93e1
\end{thebibliography}

>>>>

% ## to-do items
% --------------
% - test the cmb dipole; cite Secrest et al. NO -- ABBY's PAPER -- BUT DISCUSS THIS POINT.
% - explain, justify, or change how DD, DR, RD, and RR are counted -- the asymmetry between the tree and the sample compared to the tree.
% - discuss more philosophically why alms are sensitive tests of anisotropy
% - make some figure that quantitatively checks Gaia calibration
% - put the expectation hatN_DD onto the non-divided cumulative number plot
% - put in all missing references
% - address all CITE, XXX, YYY, ZZZ, HOGG, KSF in the text
% - make sure the figure axis labels and figure captions use precisely the math symbols used in the text; maybe also turn on TeX in matplotlib?

@@

\title{\large%
Testing cosmic isotropy and homogeneity \\ with the \textsl{Quaia} all-sky spectroscopic sample of quasars}
\author[0000-0003-2866-9403]{David W. Hogg}
\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University}
\affil{Max-Planck-Institut f\"ur Astronomie}
\affil{Flatiron Institute, a division of the Simons Foundation}

@@

\begin{abstract}\noindent
All cosmological models in general relativity are built on the assumption of large-scale isotropy and homogeneity.
These assumptions are testable.
Tests of cosmic isotropy and homogeneity are also strong tests of the quality of data or catalogs and their calibration.
Test precision increases as observations cover more solid angle, become more sensitive to distant objects, and are made with more uniform data.
The ESA \textsl{Gaia} Mission is primarily concerned with stars in the Milky Way, but it also observes millions of quasars, all sky (modulo Galactic extinction), to redshifts of around $4$.
It has produced the largest-comoving-volume public quasar catalog to date.
Here we use the $755\,850$ quasars in the \textsl{Quaia} $G<20.0$ all-sky quasar sample, built from \textsl{Gaia} and \textsl{unWISE} catalogs, to show that the angular distribution, redshift distribution as a function of hemisphere, amplitude of clustering as a function of sky position, and counts of pairs as a function of separation (fractal dimension), are all consistent with large-scale isotropy and homogeneity at the precision possible given the size of the catalog and the amplitude of the large-scale structure.
Roughly speaking, the results can be summarized by saying that they show the Universe to be isotropic at the sub-percent level and homogeneous in the sense of having a fractal dimension of $3.002\pm 0.002$.
There is no sign of any North--South density or power asymmetry.
\abby{Is this the same hemispherical power asymmetry claimed from the CMB?}
\hogg{No, unfortunately; I will do that split too.}
The isotropy results can also be interpreted as a strong test of the spectrophotometric uniformity of the \textsl{Quaia} Catalog.
\end{abstract}

\keywords{%
calibration
---
catalogs
---
cosmic~isotropy
---
cosmological~principle
---
cosmology
---
large-scale~structure~of~the~Universe
---
quasars
}
\section*{}
\clearpage
\section{Introduction}\label{sec:intro}\noindent
One of the fundamental assumptions of the physical cosmological model is that the Universe is isotropic and homogeneous on large scales (for example, \citealt{peebles, ryden}).
This assumption is testable.
It is worthy of test, both because it is always worth testing fundamental assumptions, and because the large-scale structure has self-similar or fractal characteristics, which are characteristic of inhomogeneous distributions.
And, even in the ascendent physical model of cosmology, there is expected to be at least some power in the initial conditions on all length scale.

@@

All that said, we know now that the Universe is not a fractal on large scales (\citealt{hogg05}; CITE OTHERS).
There are other reasons to continue to test isotropy and homogeneity however.
One is that the cosmic microwave background appears to show a power asymmetry, in which there appears to be a larger amplitude of fluctuations in the North/South Ecliptic (\hogg{this is not quite right, or is it?}) Cap than there is in the South/North (\citealt{wmapanisotropy}; \citealt{planckanisotropy}).
If this asymmetry is real, it should show itself in the large-scale structure as well (\citealt{zhai}).

@@

\begin{figure}[t!]
  \begin{mdframed}
  \color{captiongray}
  \begin{center}
    \includegraphics[width=\figurewidth]{notebooks/cumulativeDD_DR.png}
  \end{center}
    \caption{\textsl{Top:} The cumulative total number $N_{DD}$ \eqref{eq:DD} of quasar--quasar pairs as a function of comoving separation $s$ for the NGC and SGC subsamples. Jackknife uncertainty estimates are shown but are smaller than the points for all points.
    \textsl{Bottom:} The same, but divided by the expectation $\hat{N}_{DD}$ \eqref{eq:DDhat} in a homogeneous universe.
    This figure demonstrates that the fractal dimension is extremely close to $3$.
    \hogg{Explain the fractal-dimension label.}\label{fig:cumulative}}
  \end{mdframed}
\end{figure}
The cumulative number $N_{DD}$ of pairs up to separation $s$ is shown as a function of $s$ in the top panel of \figref{fig:cumulative}.
The top panel shows the raw counts of pairs, and the bottom panel shows the pairs relative to an expectation derived from the random catalog, which (by construction) has fractal dimension $d=3$.
Although the scaling in the top panel of \figref{fig:cumulative} looks very good, it can be biased by the finite volume within which we have observations; hence the ratios plotted in the bottom panel of \figref{fig:cumulative}.
In detail, the statistics that are plotted are cumulative pair counts obtained as follows:
We start by counting pairs within the quasar sample, between the quasar sample and the random catalog, and within the random catalog, by
\begin{align}\label{eq:DD}
    N_{DD}(s) &= \sum_n \sum_{n'\ne n} I(r_{nn'} < s) \\
    N_{DR}(s) &= f\,\sum_n \sum_{m'} I(r_{nm'} < s) \\
    N_{RD}(s) &= f\,\sum_m \sum_{n'} I(r_{mn'} < s) \\
    N_{RR}(s) &= f^2\,\sum_m \sum_{m'\ne m} I(r_{mm'} < s)\label{eq:RR} ~,
\end{align}
where the sums over $n$ or $n'$ are over quasars,
the sums over $m$ or $m'$ are over random points,
$r_{nn'}$ is the comoving distance between quasar $n$ and quasar $n'$,
$I(r<s)$ is an indicator function that returns 1 if $r<s$ and 0 otherwise,
and $f$ is a dimensionless ratio of the density of the quasar sample (on the sky) to the density of the random catalog (on the sky).
With these counts in hand, the expectation $\hat{N}_{DD}(s)$ we use for the cumulative pair counts $N_{DD}(s)$ is given by
\begin{equation}\label{eq:DDhat}
    \hat{N}_{DD}(s) = N_{DR}(s) + N_{RD}(s) - N_{RR}(s) ~,
\end{equation}
which is inspired by the currently standard correlation-function estimator (\citealt{ls}).

\hogg{\figref{fig:cumulative} strongly suggests that the fractal dimension $D$ is 3.
Our best-fit value is ... WHAT?
Get quantitative here and explain how we computed the fractal dimension.}

@@

\begin{thebibliography}{dummy}
\bibitem[Astropy Collaboration et al.(2022)]{astropy} Astropy Collaboration, Price-Whelan, A.~M., Lim, P.~L., et al.\ 2022, \apj, 935, 167. doi:10.3847/1538-4357/ac7c74
\bibitem[Eisenstein et al.(2005)]{baf} Eisenstein, D.~J., Zehavi, I., Hogg, D.~W., et al.\ 2005, \apj, 633, 560. doi:10.1086/466512
\bibitem[Eriksen et al.(2004)]{wmapanisotropy} Eriksen, H.~K., Hansen, F.~K., Banday, A.~J., et al.\ 2004, \apj, 605, 14. doi:10.1086/382267
\bibitem[Harris et al.(2020)]{numpy} Harris,~C.~R., Millman,~K.~J., van~der~Walt,~S.~J., et al.\ 2020, \nat, 585, 357. doi:10.1038/s41586-020-2649-2
\bibitem[Hoffleit \& Jaschek(1991)]{bsc}
Hoffleit, D. \& Jaschek, C.\ 1991, The Bright Star Catalog. New Haven: Yale University Observatory, 5th rev. ed.
\bibitem[Hogg et al.(2005)]{hogg05} Hogg, D.~W., Eisenstein, D.~J., Blanton, M.~R., et al.\ 2005, \apj, 624, 54. doi:10.1086/429084
\bibitem[Landy \& Szalay(1993)]{ls} Landy, S.~D. \& Szalay, A.~S.\ 1993, \apj, 412, 64. doi:10.1086/172900
\bibitem[Peebles(1993)]{peebles} Peebles, P.~J.~E. 1993, Principles of physical cosmology, Princeton University Press.
\bibitem[Planck Collaboration(2014)]{planckanisotropy} Planck Collaboration, Ade, P.~A.~R., Aghanim, N., et al.\ 2014, \aap, 571, A23. doi:10.1051/0004-6361/201321534
\bibitem[Planck Collaboration(2020)]{planck18} Planck Collaboration, Aghanim, N., Akrami, Y., et al.\ 2020, \aap, 641, A6. doi:10.1051/0004-6361/201833910
\bibitem[Ryden(2017)]{ryden} Ryden, B. 2017, \textit{Introduction to cosmology}, Cambridge University Press.
\bibitem[Schlegel et al.(1998)]{sfd} Schlegel, D.~J., Finkbeiner, D.~P., \& Davis, M.\ 1998, \apj, 500, 525. doi:10.1086/305772
\bibitem[Storey-Fisher et al.(2023)]{ksf} Storey-Fisher, K, Hogg, D.~W., Eilers, A.-C., et al.\ 2023, arXiv:2306.17749.
\bibitem[Virtanen et al.(2020)]{scipy} Virtanen,~P., Gommers,~R., Oliphant,~T.~E., et al.\ 2020, Nature Methods, 17, 261. doi:10.1038/s41592-019-0686-2
\bibitem[Zhai \& Blanton(2017)]{zhai} Zhai, Z. \& Blanton, M.~R.\ 2017, \apj, 850, 41. doi:10.3847/1538-4357/aa93e1
\end{thebibliography}
